---
title: "Age Prediction Using Multi-Omic Data"
author: "Dr. R. Craig Stillwell"
output: github_document
---

## Introduction

This report analyzes gene expression data to uncover patterns, relationships, and predictive insights. The analysis covers data preparation, correlation analysis, dimensionality reduction (PCA), clustering, and predictive modeling using machine learning techniques like Random Forest.

```{r setup, include=FALSE}
# Load necessary libraries
library(data.table)    # Efficient data handling
library(ggplot2)       # Data visualization
library(plotly)        # Interactive plots
library(corrplot)      # Correlation plots
library(factoextra)    # Visualization for PCA and clustering
library(randomForest)  # Random Forest model
library(pheatmap)      # Heatmap visualization
```

## Data Preparation

We start by loading the dataset and checking for missing values, ensuring data quality before analysis.

```{r data-loading, echo=TRUE}
# Load the dataset (adjust the filename if necessary)
data <- fread("data/test_rows.csv")

# Preview the data
head(data)

# Check the structure of the dataset
str(data)

# Check for missing values
missing_values <- sapply(data, function(x) sum(is.na(x)))
missing_values[missing_values > 0]
```

## Correlation Analysis

We compute and visualize the correlation matrix to understand relationships between genes.

```{r correlation-matrix, echo=TRUE}
# Compute the correlation matrix
correlation_matrix <- cor(data, use = "complete.obs")

# Adjust margins to give more space for the title and labels
par(mar = c(5, 5, 4.7, 5)) # Bottom, left, top, right margins

# Visualize the correlation matrix
corrplot(correlation_matrix, method = "color", tl.col = "black", tl.cex = 0.7)

# Add a properly spaced title
title(main = "Gene Expression Correlation Matrix", line = 4, cex.main = 1.5)
``` 

## Distribution Analysis

Analyzing the distribution of gene expression levels helps identify skewness or outliers in the data.

```{r gene-distribution, echo=TRUE}
# Plot distributions of RPA2_3 expression
ggplot(data, aes(x = RPA2_3)) +
  geom_histogram(binwidth = 0.5, fill = "blue", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Distribution of RPA2_3 Expression", x = "RPA2_3 Expression Level", y = "Frequency")
```

## Principal Component Analysis (PCA)

PCA reduces dimensionality and highlights the main components contributing to variation in gene expression.

```{r pca-analysis, echo=TRUE}
# Perform PCA on the dataset
pca_result <- prcomp(data, scale. = TRUE)

# Visualize explained variance
fviz_eig(pca_result, addlabels = TRUE, ylim = c(0, 50), main = "Scree Plot of Principal Components")

# Plot the PCA biplot
fviz_pca_biplot(pca_result, repel = TRUE, col.var = "contrib", 
                gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                title = "PCA Biplot of Gene Expression Data")
```

## Clustering Analysis

Clustering helps group similar samples based on gene expression patterns.

```{r kmeans-clustering, echo=TRUE}
# Elbow method to determine optimal clusters
set.seed(123)
wss <- sapply(1:10, function(k) kmeans(data, centers = k, nstart = 20, iter.max = 300)$tot.withinss)

# Plot Elbow curve
plot(1:10, wss, type = "b", pch = 19, frame = FALSE,
     xlab = "Number of clusters K", ylab = "Total within-clusters sum of squares",
     main = "Elbow Method for Determining Optimal K")

# Apply K-means clustering with K=3
kmeans_result <- kmeans(data, centers = 3, nstart = 20)

# Add cluster labels to data
data$cluster <- as.factor(kmeans_result$cluster)

# Visualize clusters
ggplot(data, aes(x = RPA2_3, y = ZYG11A_4, color = cluster)) +
  geom_point(alpha = 0.7) +
  theme_minimal() +
  labs(title = "K-Means Clustering of Gene Expression Data", x = "RPA2_3 Expression", y = "ZYG11A_4 Expression")
``` 

## Predictive Modeling

Random Forest is a powerful machine learning technique that can handle high-dimensional data and capture complex relationships. Here, we use it to predict individual age based on gene expression data.

```{r random-forest, echo=TRUE}
# Create a Synthetic 'Age' Column for testing
data$Age <- sample(20:80, nrow(data), replace = TRUE)  # Random age values between 20 and 80
# Train-test split (80/20 split)
set.seed(42)
train_idx <- sample(seq_len(nrow(data)), size = 0.8 * nrow(data))
train_data <- data[train_idx, ]
test_data <- data[-train_idx, ]

# Train Random Forest model
rf_model <- randomForest(Age ~ ., data = train_data, proximity = TRUE, importance = TRUE, ntree = 500)

# Predict on test set
predictions <- predict(rf_model, newdata = test_data)

# Evaluate model performance
actual <- test_data$Age
mse <- mean((predictions - actual)^2)
cat("Mean Squared Error (MSE):", mse)

# Visualize feature importance
varImpPlot(rf_model, main = "Feature Importance - Gene Expression Data")
```

## Heatmap of Gene Correlations

Heatmaps provide a visual summary of correlations among genes, highlighting clusters and patterns.

```{r heatmap, echo=TRUE}
# Remove cluster column for correlation analysis
data_numeric <- data[, !("cluster"), with = FALSE]

# Compute the correlation matrix
gene_correlation <- cor(data_numeric, use = "complete.obs")

# Create heatmap
pheatmap(gene_correlation, color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
         main = "Heatmap of Gene Expression Correlations", display_numbers = TRUE)
```

## Conclusion

This analysis provides insights into gene expression patterns, clustering, and feature importance. By leveraging machine learning techniques such as PCA and Random Forest, we can identify key features and underlying structures in the data, aiding future biological interpretations and research directions.